---
title: "Basic CGM Statistics"
author: "Andy Choens"
output: html_document

---

# Objectives

1. Load CGM data.
2. Clean-Up / Process the data (Data Munging!) to prepare it for use.
3. Use R to produce descriptive tables which are clinically similar to
   three of the data elements provided in the Medtronic Statistics
   table.
4. Learn more about Karen's average blood glucose levels and use this
   to discuss the diabetes management goals of this project.

All code will be shown and documented.

```{r init, echo=TRUE, message=FALSE}

## Dependencies ----------------------------------------------------------------
## If any of these fail, run the following (capitalization counts):
## install.packages("name_of_package")
library(devtools)
library(dplyr)
library(ggplot2)
library(pander)
load_all()

## Config ----------------------------------------------------------------------

## Prevents pander from wrapping the table @ 80 chars.
panderOptions('table.split.table', Inf)

```

# Data Import & Munging

The following code chunk imports the most recent CSV file available in
the data folder and stores it in a data frame called "entries". This
name was chosen for consistency with Nightscout data structures.

```{r import-data, echo=TRUE}

## Import data------------------------------------------------------------------
data(entries)

## Data Munging ----------------------------------------------------------------

## We only want the SVG Rows
entries <- entries %>%
    filter(type == "sgv")

## Week Labels - Useful labels we will be to aggregate by later.
## Does not alter all rows.
entries$wk_lbl <- NA
entries$wk_lbl[ entries$date >= '2015-06-28' & entries$date <'2015-07-05' ] <- 'Week 1: June 28 - July 04'
entries$wk_lbl[ entries$date >= '2015-07-05' & entries$date <'2015-07-12' ] <- 'Week 2: July 05 - July 11'
entries$wk_lbl[ entries$date >= '2015-07-12' & entries$date <'2015-07-19' ] <- 'Week 3: July 12 - July 18'
entries$wk_lbl[ entries$date >= '2015-07-19' & entries$date <'2015-07-26' ] <- 'Week 4: July 19 - July 25'

```

Nightscout data is a time-series data set. The descriptive statistics
shown below aggregates CGM readings by week. The "Week Labels" column
(wk\_lbl) is a simple way to aggregate, order and label the
data. Future posts will demonstrate how to graph individual CGM
readings to look for patterns. Eventually, I will demonstrate more
formal time series analysis methods.

# Descriptive Statistics

These are the measures of interest found in the Medtronic Statistics
table. The Replicated column identifies which measures are replicated
more or less exactly and which are clinically similar, due to the
differences in source data as previously discussed.

Data Element|Statistics Table     |Original Data Source              | Replicated        |Data Element Description
:----------:|:--------------------|:---------------------------------|:-----------------:|:------------------------------------------------------------------------------------------------
1           |Avg BG (mg/dL)       |Manual Blood Glucose Test         |Clinically Similar |Average blood glucose levels and standard deviation
2           |BG Readings          |Manual Blood Glucose Test         |Clinically Similar |Number of manual tests during measurement period and the average number of manual tests per a day
3           |Readings Above Target|Manual Blood Glucose Test         |Clinically Similar |Number and proportion of CGM readings above 140 mg/dL
4           |Readings Below Target|Manual Blood Glucose Test         |Clinically Similar |Number and proportion of CGM readings below 70 mg/dL
5           |Sensor Avg (mg/dL)   |Continuous Glucose Monitor Reading|Yes                |Average CGM reading during the measurement period

In this table, Medtronic uses manual blood glucose tests far more
frequently than it uses CGM data. I don't know why, but the graphics
in the report, which we will discuss soon, are all based on CGM
data. This is the only place in the report which is so heavily biased
toward blood glucose readings. Perhaps this serves as a point of
comparison between the two methodologies, to allow users to compare
results.

# Sensor Avg (mg/dL)

The table shown below is calculated in the same way "Sensor Avg" is
one row five. This measure is clinically similar to Avg BG measure on
row 1. In either case, it is a simple measure of blood glucose
control. Because it is an average, hypoglycemic episodes can
mathematically cancel out periods of hyperglycemia. Unfortunately,
they do not cancel out damage caused by hyperglycemia.

To use this measure to understand diabetes management, you must use
both the average and the standard deviation, to account for the
variance in blood glucose levels. Diabetics who swing to extreme blood
glucose levels quickly, such as Karen, have high standard deviations,
even if the average is relatively close to 100.


```{r avg-bg, echo=TRUE}

entries %>%
    filter( !is.na(wk_lbl) ) %>%
    group_by( "Date Range" = wk_lbl) %>%
    summarise("Avg Sensor Reading" = paste(round(mean(sgv, na.rm=TRUE),1),"mg/dL"),
              "Std Deviation" = paste( format( round(sd(sgv, na.rm=TRUE),1),nsmall=1),"mg/dL")
              ) %>%
    pander()

``` 

The blood glucose level for a non-diabetic is, on average, 100. Over
the course of the day, a non-diabetic could be as high as 140 and as
low as 70, although such extremes values would be short-lived. This
shows us that even with insulin treatment, Karen's averages are higher
than a non-diabetic. The weekly averages shown above are a little
higher than the report from February, but well within the range we
would expect to see based on the variance seen across weeks. The high
standard deviation demonstrates that Karen experiences a wide range of
blood glucose levels. Although she does not meet the definition of a
"brittle diabetic", her blood glucose levels do swing dramatically as
a result of her
[gastroparesis](https://en.wikipedia.org/wiki/Gastroparesis), a
complication of diabetes.

# Number of Sensor Readings

The Statistics table does not include the number of sensor
readings. It only includes the number of manual blood glucose
readings, which cannot be directly replicated using CGM data. It is
surprising that the Medtronic report only includes the number and
daily average of manual tests. Knowing the number of sensor readings
is useful for understanding the consistency with which the CGM is
worn, which is obviously important for assessing the accuracy and
validity of the aggregate CGM data.

This measure is not really clinically similar to Measure 2. Manual
blood glucose tests are important. They are used to calibrate the CGM
sensor, which helps improve the accuracy of the CGM data we are going
to rely on so much here. Reporting on the number of sensor readings is
as similar as we can get, but it is not clinically similar. But, as a
measure, the count and distribution of CGM sensor readings is worth
discussing.

When the CGM is worn correctly and is operational, it records a new
reading every five minutes. But, Murphy rules. Batteries die, sensors
fall out and sensors lose their connection to the rig. This is an
individualized "big" data set and it isn't even genetic! In an ideal
world, the sensor will record 2,016 readings every week. As you can
see, the number of actual readings is usually below 2,000. When the
number of readings falls unexpectedly, such as the week starting on
July 19, it may indicate a problem with the CGM or how consistently
Karen is using her CGM.

```{r n-sensor-readings, echo=TRUE}

entries %>%
    filter( !is.na(wk_lbl) ) %>%
    group_by( "Date Range" = wk_lbl) %>%
    summarize("N Entries" = format( n(), big.mark="," ) ) %>%
    pander()

```

In this case, the low number of CGM readings is misleading. It is not
indicative of a CGM error. Not is it telling us that Karen decided to
stop managing her diabetes for a week. We were in Ireland that week on
vacation. Contrary to what you see, Karen did use her CGM in Europe.
There are gaps in the Nightscout data because we stopped using the rig
for several days that week. We would like to thank Verizon for making
it outrageously complicated and expensive to set up and monitor a data
connection in Europe. THANK YOU!

# Time Spent Where

Duration matters. The Statistics table includes several measures to
approximate how much time is spent in different blood glucose value
ranges. Karen's endocrinologist wants to know how much time she spends
in the "good" range, 70-140 mg/dL. This range of values damages her
body less than either hyper / hypo-glycemia. In the Medtronic report,
these data elements are based on manual blood glucose tests, but we
can create a clinically similar measure using Nightscout data. The
following table shows the percentage of CGM readings that are in the
target range, above it and below it. A high proportion of readings in
the 70 - 140 mg/dL range is the desired outcome.

```{r ranges-bg, echo=TRUE}

entries %>%
    filter( !is.na(wk_lbl) ) %>%
    group_by( "Date Range" = wk_lbl) %>%
    summarise("Sensor Readings" = n(),
              "Readings In Target Range " = paste( format(round(sum(ifelse(sgv >= 70 & sgv<= 140,1,0))/n()*100,2),
                                                          nsmall = 2),
                                                   "%", sep=""),
              "Readings Above Target" = paste( format(round(sum(ifelse(sgv > 140,1,0))/n()*100,2),
                                              nsmall = 2),
                                              "%",sep=""),
              "Readings Below Target" = paste( format(round(sum(ifelse(sgv < 70,1,0))/n()*100,2),
                                               nsmall=2),
                                               "%",sep="")
              ) %>%
    pander()

```

Although mathematically crude, simple proportions such as these are
similar to the AUC and it is easier for many patients and doctors to
understand. By including the average sensor reading for readings in a
given range we can better understand the interplay between duration
and severity. 

## Readings In Target Range

The first column, the measure denominator, is the total number of CGM
readings. The numerator, the number of readings in the target range
(70-140 mg/dL) is the second column. A diabetic wants as many readings
as possible to fall in this range. The last column is the mean sensor
value for readings in the target group. An average close to 100 mg/dL
is good.

```{r on-target, echo=TRUE}

entries %>%
    filter( !is.na(wk_lbl) ) %>%
    group_by( "Date Range" = wk_lbl) %>%
    summarise("CGM Readings" = n(),
              "Readings In Target Range" = sum(ifelse(sgv >= 70 & sgv<= 140,1,0)),
              "% In Target Range" = format(round(sum(ifelse(sgv >= 70 & sgv<= 140,1,0))/n()*100,2),
                                                  nsmall = 2
                                                  ),
              "CGM AVG" = paste( round(mean( ifelse(sgv >= 70 & sgv <= 140, sgv, NA), na.rm=TRUE),2),
                                "mg/dL",
                                sep=" "
                               )
              ) %>%
    pander()

```

## Readings Above Target Range

The first column, the measure denominator, is the total number of CGM
readings. The second column, the numerator, is the number of readings
above 140 mg/dL. A diabetic wants to minimize the number of readings
above the target range. The last column is the mean sensor value for
readings above the target range. Time spent above 140 mg/dL is
damaging to the body. At extreme values, hyperglycemia can result in
diabetic ketoacidosis, coma and death. A hyperglycemic average close
to 140 is better than a higher average. Reducing the number of
readings about 140 is ALWAYS a good thing.

```{r above-target, echo=TRUE}

entries %>%
    filter( !is.na(wk_lbl) ) %>%
    group_by( "Date Range" = wk_lbl) %>%
    summarise("CGM Readings" = n(),
              "Readings Above Target" = sum(ifelse(sgv > 140,1,0)),
              "% Above Target" = paste(format(round(sum(ifelse(sgv > 140,1,0))/n()*100,2),
                                              nsmall = 2),
                                       "%"),
              "CGM AVG" = paste( format(round(mean( ifelse(sgv > 140, sgv, NA), na.rm=TRUE),2),nsmall=2), " mg/dL")
              ) %>%
    pander()

```

## Readings Below Target Range

The first column, the measure denominator, is the total number of CGM
readings. The second column, the numerator, is the number of readings
below 70 mg/dL. A diabetic wants to minimize the number of readings
below the target range. When below 70 mg/dL diabetics experience
confusion, anxiety and other physiological and psychological
symptoms. Severe lows can result in losing consciousness or diabetic
comas. Thankfully, the latter is rare. The last column is the mean CGM
value for readings below the target range. A hypoglycemic average
close to 70 is better than a lower average.

```{r below-target, echo=TRUE}

entries %>%
    filter( !is.na(wk_lbl) ) %>%
    group_by( "Date Range" = wk_lbl) %>%
    summarise("CGM Readings" = n(),
              "Readings Below Target" = sum(ifelse(sgv < 70,1,0)),
              "% Below Target" = format(round(sum(ifelse(sgv < 70,1,0))/n()*100,2),
                                                  nsmall = 2
                                                  ),
              "BG AVG" = paste( format( round( mean( ifelse(sgv < 70, sgv, NA), na.rm=TRUE), 2),
                                        nsmall=2
                                      ),
                                "mg/dL",
                                sep=" "
                              )
              ) %>%
    pander()

```
